Towards an Ecological Trait-data Standard

### 2018-05-28 manuscript/manuscript.Rmd
update manuscript: comments of MG

Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07]. Because trait-based approaches allow studying both patterns and mechanisms [@lavorel02; @diaz16], recent years have seen a proliferation of trait-based research in a wide range of fields.
Trait-based studies have been conducted in a wide range of thematic areas ranging from the evolutionary basis of individual-level properties [@salguero-gomez16] to global patterns of biodiversity [@diaz16] and ecosystem functioning [@allan15; @bello10]. With this mechanistic depth, the trait framework relates losses of ecosystem functions to changes in biodiversity, and on to climate change or local anthropogenic land use [@diaz11; @lavorel12; @allan15]. <!-- ;@wood15;  @diaz11; @jennings09; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15-->
Using traits is also a promising means of bypassing taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms. This is because functional traits allow us to infer the ecological role of organisms  from their apparent features, regardless of their taxonomic identity [@leprovost17; @duarte11; @schrodt15].

@@

Many issues in trait-based research arise when compiling datasets from several sources. Data may differ in taxonomic nomenclature and resolution (e.g. reported on species level or aggregated on higher taxonomical orders), the scale and place of the study context, or the methodology applied in measurements. All of these factors render trait data extremely heterogeneous and make the task of data compilation time-consuming or even prohibitive.
However, fully exploiting the potential of trait-based approaches relies heavily on the broad availability and compatibility of trait data to achieve sufficient taxonomic and regional coverage, both of present-day taxa ase well as in evolutionary deep-time.

<!-- 3. trait databases have been developed, with standard structure and terms --> 
To this end, trait data have been standardised and compiled in centralised databases for specific organism groups and regional scope, often centred around particular research questions [e.g. PanTHERIA, @jones09; TRY, @kattge11; AmphiBio, @oliveira17]. 
These databases are then published as static data tables (i.e. product of a one-time effort as it is the case of AmphiBIO), or in the form of sophisticated online portals, which attract data submissions from a focal community of researchers and provide research networks for consensus building in data resolution and measurement methods (e.g. TRY, www.try-db.org). Furthermore, these portals map heterogeneous data into a common scheme and, importantly, also offer access control and data usage policies. As such, they protect the rights of the original data providers while simplifying data queries for synthesis researchers. 
Besides initiatives aiming at assembling data, other tools to enable the compatibility of data across databases are being developed. These include semantic-web standards [@wieczorek12; @page08] and ontologies of standard terms [@walls12; @garnier17].
<!-- Solving the numerous issues of data assemblage and harmonisation in centralised infrastructures requires a huge effort in costs and personnel. -->
<!-- 4. distributed trait data must be integrated --> 
While these community efforts for centralisation and harmonization are great advancements in biodiversity informatics and data-management, the number of decentralised trait datasets is increasing rapidly. This is due to the fact that open-science reaches the mainstream: it has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12; @swan12; @allison15]. As a result, an increasing number of individual research projects comply with open-data standards and publish their primary data on file hosting services like Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org, where no data standards are forced upon the uploaded material. <!-- The data standards mentioned above, that have been proposed for relational trait-databases, do hardly apply for these individual datasets.-->
It is likely that data will become increasingly available, but a lack of data and metadata standardisation will hamper the efficient re-use and synthesis of published datasets.

@@

In the field of comparative biology, morphological traits such as plant flower, leaf and stem traits or bird wing and beak measurements, as well as life-history traits such as Ellenberg values  have been measured for decades, and have been published in regular journal articles or books. 
With the rise of ecological trait-based research, individual measurements and information available from species descriptions have been compiled into project-specific datasets that typically comprise a local set of taxa and a focal set of traits. 
A plethora of such static datasets has been published along with scientific articles or as standalone data publications [see @kleyer08 for a review on plant data; on animal data, see e.g., @gossner15; @ricklefs17].<!--  @sokol11;  @renner14; @vanderplas17; @price13; -->
Today, the online publication of such data is greatly facilitated by  *file hosting services* (e.g. Figshare.com), which warrant long-term accessibility and citability via DOIs, and support Public Domain dedication or Creative Commons licenses.
These platforms offer publicly accessible repositories at low-cost or for free, which makes them attractive for small and intermediate sized research projects that cannot dedicate extra resources for data management. 
However, although open for manual access, the trait datasets on data repositories might be stored in proprietary (e.g. .xlsx, .docx) or binary (e.g. .pdf) data formats which make a programmatical extraction tedious and dependent on commercial software, putting the long-term and open accessibility of these data at risk. 
Most importantly, these platforms enable public hosting of data with very low thresholds for *metadata* documentation and data standardisation. 
<!-- As a result, datasets greatly vary in structure and terminology. -->

For trait data, there are typical issues arising from the variability of data structures. 
For instance, the column descriptions and terminology applied to taxa and traits are project specific, and rarely chosen to allow translation into larger database initiatives. Furthermore, metadata varies in its detail, e.g. for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a]. 
In terms of structure, trait data usually are reported in a species$\times$traits wide-table format. In this format, each row contains a species (or taxon) for which multiple traits are reported in columns. 
Similarly, when reporting raw data, researchers place observations of individual organisms in rows with multiple trait measurements applied to the same individual across multiple columns. 
Variability in the number and meaning of columns in these *data tables* requires tedious manual adjustments when merging multiple datasets [@wickham14].

@@

<!-- database backend, selective download, user access management --> 
More specialised trait-database platforms have been created to cover certain trait types (e.g. floral traits, seed traits, root traits or wood density traits), interaction types (e.g. pollination traits or feeding relationships), or a specific environmental and experimental context of the trait observation (e.g. location or climatic data).
Such database initiatives attract data submissions from a defined research field and take care of the harmonisation process and thereby greatly facilitate data synthesis.
<!--For example, traits on crop plants have been collected in agronomy databases like  BetyDB [https://www.betydb.org/; @lebauer17].-->
For example, by aiming for a universal framework for plant traits, the TRY database for plant traits [@kattge11]  attracted more data submissions and downloads than any other trait data platform. The *online database* enables selective data download and user permission and rights management. As a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section). 
Microbial ecologists also make frequent use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14; @fierer14;  @fierer12]. Here, taxa are derived from metagenomic analysis [@torsvik02; @langille13]. Databases are also used to interpret gene sequences, so called 'operational taxonomic units' (OTUs), in terms of their functional role [e.g. the KEGG orthology, @kanehisa12]. 
For animals, a single unified platform and harmonising scheme for animal trait data is still lacking. The reason for this may be that harmonising trait data on animals, which span multiple trophic levels and possess diverse body plans, is a more complex task than for plants [@moretti17]. Nonetheless, the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/).

@@

Assigning an unambiguous methodological protocol to a trait, including the units to use or the ordinal or factor levels to be assigned, is key for standardising the physical process of measuring. Efforts to develop handbooks for measurement protocols provide such a methodological standardisation for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17], but obviously are of limited use in harmonising trait data that predate or ignore this standard [@kattge11a].
Very general classes of traits are defined within the list of GeoBON Essential Biodiversity Variables [@pereira13].

@@

While trait thesauri and ontologies typically define traits for focal groups of organisms, they do not specify the  format or structure in which trait data should be stored and linked to further standard terminologies, e.g. for taxa. 
To make sense of trait data in the context of more general databases, a consensus definition of trait data is necessary.

Trait data have been defined by @garnier17 to follow an entity-quality model (EQ), where a trait observation is 'an entity having a quality'. 
More specifically, a trait dataset contains information on quantitative *measurements* or qualitative *facts* (i.e. trait values) describing the physical phenotypic characteristics relating to fitness, behaviour or performance (i.e. traits) observed on a biological entity (i.e. an individual specimen, or parts of an individual specimen) that can be assigned to a biological taxon (i.e. a species or higher-level taxon). 
We are expanding on this definition: quantitative measurements are values obtained either by direct morphological, physiological or behavioural observations on single specimens (Fig. 1*a*), by aggregating replicated measurements on multiple entities (Fig. 1*b*) or by estimating the means or ranges for the respective taxon as reported in the literature or other published sources (e.g. databases, Fig. 1*c*). Qualitative facts are assignments of an entity to a categorical level, e.g. of a behavioural or life-history trait (Fig. 1*d*). 
The entity or observation (i.e. the *occurrence*) to which the reported measurement or fact applies may differ in organisational scale -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or a higher-level taxon (e.g. a genus).

\begin{figure}[t]
\centering
\includegraphics[width=0.85\textwidth, angle=0]{figures/fig1.png}
\caption{Types of ecological trait data assume different entities or reported quantities. a) morphometric or morphological measurements of individual body features (lengths, areas, volumes, weights) or other quantities related to life history (e.g. reproductive rates, life spans); b) aggregated traits are reported as means taken on multiple measures of members of a taxon; c) quantities may be extracted from literature or existing databases, referring to the entire taxon (or a subset, e.g. a sex) as the entity of description; d) Qualitative traits are categorical or binary descriptors of the entire taxon.}

@@

It appears to be broad consensus that an open biodiversity science is crucial for an evidence-based decision making and conservation policy on regional and global scales.
In times of increasing demand for open research data and international platforms for biodiversity data management, the development of meaningful terminologies for the standardisation of biodiversity data is more than essential: defined ontologies enable researchers to relate published datasets to each other to achieve a greater synthesis, thereby paving the way for new insights on global biodiversity patterns. In terms of data science, machine-readable, ontology-based data ease the application of big-data mining and machine-learning techniques.

>>>>

Functional traits are phenotypic  (i.e. morphological, physiological, behavioral) characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07]. Because trait-based approaches allow studying both patterns and mechanisms [@lavorel02; @diaz16], recent years have seen a proliferation of trait-based research in a wide range of fields.
Trait-based studies have been conducted in a wide range of thematic areas ranging from the evolutionary basis of individual-level properties [@salguero-gomez16] to global patterns of biodiversity [@diaz16] and ecosystem functioning [@allan15; @bello10]. 
The trait framework relates losses of ecosystem function to changes in  the functional composition of species assemblages [@mouillot13; @perovic15]. This offers the mechanistic background to relate biodiversity to climate change or local anthropogenic land use [@diaz11; @lavorel12; @allan15]. <!-- ;@wood15;  @diaz11; @jennings09; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15-->
Using traits is also a promising means of bypassing taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms. This is because functional traits allow us to infer the ecological role of organisms  from their apparent features, regardless of their taxonomic identity [@leprovost17; @duarte11; @schrodt15].

@@

Many issues in trait-based research arise when compiling datasets from several sources. Data may differ in taxonomic nomenclature and resolution (e.g. reported on species level or aggregated on higher taxonomical orders), the scale and place of the study context, or the accurracy of the methodology applied in measurements. These differences are not always documented in the metadata accompanying a dataset. 
All of these factors render trait data extremely heterogeneous and make the task of data compilation time-consuming or even prohibitive.
However, fully exploiting the potential of trait-based approaches relies heavily on the broad availability and compatibility of trait data to achieve sufficient taxonomic and regional coverage, both of present-day taxa ase well as in evolutionary deep-time.

<!-- 3. trait databases have been developed, with standard structure and terms --> 
To this end, the number of available trait datasets is increasing rapidly. 
In the past, trait data have been standardised and compiled in centralised databases for specific organism groups and regional scope, often centred around particular research questions [e.g. PanTHERIA, @jones09; TRY, @kattge11; AmphiBio, @oliveira17]. 
<!--These databases are then published as static data tables (i.e. product of a one-time effort as it is the case of AmphiBIO), or in the form of sophisticated online portals, which attract data submissions from a focal community of researchers and provide research networks for consensus building in data resolution and measurement methods (e.g. TRY, www.try-db.org). Furthermore,--> These initiatives map heterogeneous data into a common scheme and, importantly, also offer access control and data usage policies. As such, they protect the rights of the original data providers while simplifying data queries for synthesis researchers. 
Besides initiatives aiming at assembling data, other tools to enable the compatibility of data across databases are being developed. These include semantic-web standards [@wieczorek12; @page08] and ontologies of standard terms [@walls12; @garnier17].
<!-- Solving the numerous issues of data assemblage and harmonisation in centralised infrastructures requires a huge effort in costs and personnel. -->
<!-- 4. distributed trait data must be integrated --> 
Meanwhile, open-science reaches the mainstream: it has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12; @swan12; @allison15]. As a result, an increasing number of individual research projects publish their primary data on file hosting services like Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org, where no data standards are forced upon the uploaded material. <!-- The data standards mentioned above, that have been proposed for relational trait-databases, do hardly apply for these individual datasets.-->
It is likely that trait data will become increasingly available, but a lack of data and metadata standardisation will hamper the efficient re-use and synthesis of published datasets.

@@

In the field of comparative biology, morphological traits related to plant flower, leaf and stem traits or bird wing and beak measurements, as well as life-history traits such as Ellenberg values for plants or ecological parameters of animals (e.g. reproductive traits, feeding biology, dispersal or body size) have been measured for decades, and have been published in regular journal articles or books. 
With the rise of ecological trait-based research, individual measurements and information available from species descriptions have been compiled into project-specific datasets that typically comprise a local set of taxa and a focal set of traits. 
A plethora of such static datasets has been published along with scientific articles or as standalone data publications [see @kleyer08 for a review on plant data; on animal data, see e.g., @gossner15; @ricklefs17].<!--  @sokol11;  @renner14; @vanderplas17; @price13; -->
Today, the online publication of such data is greatly facilitated by  *file hosting services* (e.g. Figshare.com), which warrant long-term accessibility and citability via DOIs, and support Public Domain dedication or Creative Commons licenses.
These platforms offer publicly accessible repositories at low-cost or for free, which makes them attractive for small and intermediate sized research projects that cannot dedicate extra resources for data management. 
However, although open for manual access, the trait datasets on data repositories might be stored in proprietary (e.g. .xlsx, .docx) or binary (e.g. .pdf) data formats which make a programmatical extraction tedious and dependent on commercial software, putting the long-term and open accessibility of these data at risk. 
Most importantly, these platforms enable public hosting of data with very low thresholds for *metadata* documentation and data standardisation. 
<!-- As a result, datasets greatly vary in structure and terminology. -->

For trait data, there are typical issues arising from the variability of data structures. 
For instance, the column descriptions and terminology applied to taxa and traits are mostly project specific, and rarely chosen to allow translation into larger database initiatives. Furthermore, metadata varies in its detail, e.g. for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a]. 
In terms of structure, trait data usually are reported in a species$\times$traits wide-table format. In this format, each row contains a species (or taxon) for which multiple traits are reported in columns. 
Similarly, when reporting raw data, researchers place observations of individual organisms in rows with multiple trait measurements applied to the same individual across multiple columns. 
Variability in the number and meaning of columns in these *data tables* requires tedious manual adjustments when merging multiple datasets [@wickham14].

@@

<!-- database backend, selective download, user access management --> 
More specialised trait-database platforms have been created to cover certain trait types (e.g. floral traits, seed traits, root traits or wood density traits), interaction types (e.g. pollination traits or feeding relationships), or a specific environmental and experimental context of the trait observation (e.g. location or climatic data).
Such database initiatives attract data submissions from a defined research field and take care of the harmonisation process and thereby greatly facilitate data synthesis.
<!--For example, traits on crop plants have been collected in agronomy databases like  BetyDB [https://www.betydb.org/; @lebauer17].-->
For example, by aiming for a universal framework for plant traits, the TRY database [@kattge11]  attracted more data submissions and downloads than any other trait data platform. The *online database* enables selective data download and user permission and rights management. As a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section). 
Microbial ecologists also make frequent use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14; @fierer14;  @fierer12]. Here, 'operational taxonomic units' (OTUs) are derived from metagenomic analysis [@torsvik02; @langille13]. Databases are also used to interpret OTUs in terms of their functional role [e.g. the KEGG orthology, @kanehisa12]. 
For animals, a single unified platform and harmonising scheme for animal trait data is still lacking. The reason for this may be that harmonising trait data on animals, which span multiple trophic levels and possess diverse body plans, is a more complex task than for plants [@moretti17]. Nonetheless, initiatives for particular groups of animals, such as the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/), already exist.

@@

Very general classes of traits are defined within the list of GeoBON Essential Biodiversity Variables [@pereira13].
Assigning a more detailled and unambiguous methodological protocol to a trait, including the units to use or the ordinal or factor levels to be assigned, is key for standardising the physical process of measuring. Efforts to develop handbooks for measurement protocols provide such a methodological standardisation for plants  [@cornelissen03; @perez-harguindeguy13] or invertebrates [@moretti17], but obviously are of limited use in harmonising trait data that predate or ignore this standard [@kattge11a].

@@

While trait thesauri and ontologies typically define traits for focal groups of organisms, they do not specify the  format or structure in which trait data should be stored and linked to further standard terminologies, such as standard taxonomy nomenclatures.  
To make sense of trait data in the context of more general databases, a consensus definition of trait data is necessary.

Trait data have been defined by @garnier17 to follow an entity-quality model (EQ), where a trait observation is 'an entity having a quality'. 
More specifically, a trait dataset contains information on quantitative *measurements* or qualitative *facts* (i.e. trait values) describing the physical phenotypic characteristics relating to fitness and performance (i.e. traits) observed on a biological entity (i.e. an individual specimen, or parts of an individual specimen) that can be assigned to a biological taxon (i.e. a species or higher-level taxon). 
We are expanding on this definition: quantitative measurements are values obtained either by direct morphological, physiological or behavioural observations on single specimens (Fig. 1*a*), by aggregating replicated measurements on multiple entities (Fig. 1*b*) or by estimating the means or ranges for the respective taxon as reported in the literature or other published sources (e.g. databases, Fig. 1*c*). Qualitative facts are assignments of an entity to a categorical level, e.g. of a behavioural or life-history trait (Fig. 1*d*). 
The entity or observation (i.e. the *occurrence*) to which the reported measurement or fact applies may differ in organisational scale -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or a higher-level taxon (e.g. a genus).

\begin{figure}[t]
\centering
\includegraphics[width=0.85\textwidth, angle=0]{figures/fig1.png}
\caption{Types of ecological trait data assume different entities or reported quantities. a) morphometric or morphological measurements of individual body features (lengths, areas, volumes, weights) or other quantities related to life history (e.g. reproductive rates, life spans); b) aggregated traits are reported as means taken on multiple measures of members of a taxon; c) quantities may be extracted from literature or existing databases, referring to the entire taxon (or a subset, e.g. a sex) as the entity of description; d) Qualitative traits are categorical or binary descriptors of the entire species or higher taxonomic group.}

@@

It appears to be broad consensus that an open biodiversity science is crucial for an evidence-based decision making and conservation policy on regional and global scales.
In times of increasing demand for open research data and international platforms for biodiversity data management, the development of meaningful terminologies for the standardisation of biodiversity data is more than essential: defined ontologies enable researchers to relate published datasets to each other to achieve a greater synthesis, thereby paving the way for a better mechanistic understanding of the relationship between drivers, communities and functions and providing new insights on global biodiversity patterns. 
Moreover it might be also a step towards a more predictive ecology as a broader set of available traits might enable more hypothesis based trait-based approaches. 
In terms of data science, machine-readable, ontology-based data ease the application of big-data mining and machine-learning techniques.
